---
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: apa6.tex

classoption: man
lang: american
lineno: true

title: "Example manuscript demonstrating the use of the papaja template"
shorttitle: Using papaja
author: "Frederik Aust"
affiliation: University of Cologne
note: The papaja-template, helper functions and further instructions can be retrieved from https://github.com/crsh/papaja.

abstract: "This example manuscript demonstrates how to use RStudio and RMarkdown to produce an APA conform manuscript. Using pandoc your RMarkdown can be converted to HTML, PDF, or Word documents. At this point, only PDF documents adhere to the APA mansucript guidelines."

bibliography: references.bib
csl: apa6.csl
---

# What is this?
As you may have heard, recently, there has been a growing interest in reproducible research. Reproducible data analysis is an easy to implement and important aspect of the strive towards reproducibility. For *R* users, RMarkdown has been suggested as one possible framework for reproducible analyses. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. This example document assumes you have hoped onto the band wagon and know how to use RMarkdown to conduct and comment your analyses. If you're new to [RMarkdown](http://rmarkdown.rstudio.com/), I recommend you get to grips with it first.

I use [RStudio](http://www.rstudio.com/) (which makes use of [pandoc](http://johnmacfarlane.net/pandoc/)) to create my documents, but the general process should work when using pandoc directly from the command line.

# How do I use this?
With the papaja-template, when you click RStudio's *Knit* button an APA conform manuscript will be generated that includes both your text and the output of any embedded R code chunks within the manuscript.

## Printing R output
Any output from R is included as you usually would using RMarkdown.

```{r}
summary(cars)
```

For prettier tables, I suggest you have a look at my helper function `apa.table()` in the [papaja repository](https://github.com/crsh/papaja). Of course, e.g, the popular `xtable` package can also be used to create tables. Unfortunately, `xtable()` captions are [set to the left page margin](http://tex.stackexchange.com/questions/42209/centering-tables-in-document-class-apa6) (for an example, see last page of this document). `apa.table()` fixes this problem. As required by the APA guidelines, tables are on the final pages of the manuscript.

```{r results = 'asis', warning = FALSE, message = FALSE}
library("xtable")
print(
  xtable(summary(cars), caption = "Prettier table created using xtable.
         Note the caption is set to the left page margin
         instead of aligning wit the table. :(")
  , comment = FALSE
  , booktabs = TRUE
  , caption.placement = "top"
  , include.rownames = FALSE
)
```

You can also embed plots, for example:

```{r fig.cap = "Exmple figure created by in-document R code."}
plot(cars)
```

As required by the APA guidelines, figures, too, are printed to the final pages of the document.

## Citations
You can insert citations like this:

`[e.g., @bauer_2014; @bem_2011]` &#8594; [e.g., @bauer_2014; @bem_2011].

Citing without parentheses is, of course, also possible:

`@bauer_2014` &#8594; @bauer_2014.

The citation style is set in the header of this document with the `csl` parameter. The relevant references will, of course, be added to the documents references automatically. In order for citations to work, you need to supply a .bib-file to the `bibliography` parameter in the document header. See the [RMarkdown documentation](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html) and [Citation Style Language](http://citationstyles.org/) for further details.

## Document options
This text is set as manuscript. If you want a thesis-like document you can change the `classoption` in the document header from `man` to `doc`. You can also preview a polished journal typesetting by changing the `classoption` to `jou`.

Line numbering can be deactivated in by removing the `lineno` argument from the header of this document.

## Last words
That's all I have. Enjoy writing your manuscript. If you have any trouble or ideas for improvements, open an [issue](https://github.com/crsh/papaja/issues) on GitHub or make a pull request with the fix. ;)

# References
