---
title: "Example manuscript demonstrating the use of the papaja template"
shorttitle: "Using papaja"
author: 
  - name:  Frederik Aust
    affiliation: 1
    corresponding: yes    # Define only one corresponding author
    department:
    address: "Department Psychology, University of Cologne, Herbert-Lewin-Str. 2, 50931 Köln, Germany"
    email: "frederik.aust@uni-koeln.de"
affiliation:
  - id: 1
    institution: "University of Cologne"

abstract: |
  This example manuscript demonstrates how to use RStudio and RMarkdown to
  produce an APA conform manuscript. Using pandoc your RMarkdown can be converted
  to PDF or Word documents.
  
note: |  
  `papaja` has currently not been submitted to CRAN; a development version is available at https://github.com/crsh/papaja.
  
keywords: "APA, knitr, R, RMarkdown, papaja"

wordcount: "Too lazy to count"

class: man
lang: american
figsintext: yes
lineno: yes
bibliography:
  - references.bib
  - r-references.bib

output: papaja::apa6_pdf
---

```{r message = FALSE, warning = FALSE}
require("papaja")
apa_prepare_doc() # Prepare document for rendering
```

# What is papaja?
As you may have heard, recently, interest in reproducible research has been growing.
Reproducible data analysis is an easy to implement and important aspect of the strive towards reproducibility.
For *R* users, RMarkdown has been suggested as one possible framework for reproducible analyses.
`papaja` is a R-package in the making including a RMarkdown template that can be used with RStudio (or without) to produce complete manuscripts (PDF and Word documents), which conform to the American Psychological Association (APA) manuscript guidelines (6th Edition).
To do so, papaja uses the \LaTeX document class `apa6` and a `.docx`-reference file. The supplied R-functions are ment to facilitate the reporting of statistics in accordance with APA guidelines.


Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents (among others).
In the following I will assume you have hopped onto the band wagon and know how to use RMarkdown to conduct and comment your analyses.
If this is not the case, I recommend you get to grips with [RMarkdown](http://rmarkdown.rstudio.com/) first.
I use [RStudio](http://www.rstudio.com/) (which makes use of [pandoc](http://johnmacfarlane.net/pandoc/)) to create my documents, but the general process should work using any other R-editor.

# How do I use papaja?
Once you have installed `papaja` and all other [required software](https://github.com/crsh/papaja) you can select the APA template when creating a new Markdown file through the menus in RStudio.
When you click RStudio's *Knit* button, `papaja`, `rmarkdown,` and `knitr` together create an APA conform manuscript that includes both your text and the output of any embedded R code chunks within the manuscript.
Don't worry about the first chunk of R code at the beginning of the document.
It sets the stage for the creation of your document.

```{r echo = TRUE, eval = FALSE}
require("papaja")
apa_prepare_doc() # Prepare document for rendering
```

## Printing R output
Any output from R is included as you usually would using RMarkdown.
By default the R code will not be displayed in the final documents.
If you wish to show off your code you need to set `echo = TRUE` in the chunk options.

```{r echo = TRUE}
summary(cars)
```

But, surely, this is not what you want your submission to look like.
I think we can do better.

### Print tables
For prettier tables, I suggest you have a look at `apa_table()`.
Of course, e.g, the popular `xtable` or `tables` packages can also be used to create tables when knitting PDF documents.
Unfortunately, `xtable()` captions are [set to the left page margin](http://tex.stackexchange.com/questions/42209/centering-tables-in-document-class-apa6).
More importantly, these packages cannot be used when you want to create Microsoft Word documents because they rely on \LaTeX for typesetting.
`apa_table()` creates tables that conform to APA guidelines and are correctly rendered in PDF and Word documents.
But don't fool yourself.
Table formatting is somewhat limited in Word documents due to missing functionality in pandoc (e.g., it is not possible to have cells span across multiple columns).

As required by the APA guidelines, tables are pushed to the final pages of the manuscript when creating a PDF.
Again, this is not the case in Word documents due to limited pandoc functionality.
To place figures and tables in your text instead, set the `figsintext` parameter in the document header to `yes` or `true` as I have done in this document.

```{r results = 'asis', warning = FALSE, message = FALSE, echo = TRUE}
my_table <- apply(cars, 2, function(x) round(
    c(Mean = mean(x), SD = sd(x), Min = min(x), Max = max(x)), 2)
  )

apa_table(
  my_table
  , align = c("l", "r", "r")
  , caption = "A summary table of the cars dataset."
  , note = "This table was created using apa\\_table()"
  , added_colnames = "Descriptives"
)
```

The bottom line is, Word documents will be less polished than PDF.
The resulting documents should suffice to enable collaboration with Wordy colleagues and prepare a journal submission.

### Plots
You can also embed plots, for example:

```{r fig.cap = "Exmple figure created by in-document R code.", echo = TRUE}
plot(cars)
```

Again, as required by the APA guidelines, figures are pushed to the final pages of the document unless you set `figsintext` to `yes`.

### Report statistical analyses
`apa_print()` will help you report the results of your statistical analyses.
The function will format your input to produce readily reportable text.

```{r echo = TRUE}
my_regression <- lm(dist ~ speed, cars)
my_results <- apa_print(my_regression)
```

In this case speed is a significant predictor of the distance taken to stop, `r my_results$full$speed`.
The regression explains `r my_results$est$modelfit$r2` variance, which is of course statistically significant, `r my_results$stat$modelfit$r2`.

The previous paragraph was produced by the following text:

```{r eval = FALSE, echo = TRUE}
In this case speed is a significant precitor of the distance taken
to stop, `r my_results$full$speed`. The regression explains
`r my_results$est$modelfit$r2` variance, which is of course
statistically significant, `r my_results$stat$modelfit$r2`.
```

What's even more fun: You can easily create a complete regression table using by passing `my_results$table` to `apa_table()`.

```{r echo = TRUE}
apa_table(
  my_results$table
  , align = c("l", "r", "c", "r", "r", "r")
  , escape = FALSE
  , row_names = FALSE
  , caption = "Regression table"
  , note = "This is a table generated by apa\\_print() and apa\\_table()."
)
```


## Citations
You can insert citations like this:

`[e.g., @bauer_2014; @bem_2011]` &#8594; [e.g., @bauer_2014; @bem_2011].

Citing without parentheses is, of course, also possible:

`@bauer_2014` &#8594; @bauer_2014.

We have made citing R particularly easy:

```{r echo = TRUE}
my_citation <- cite_r(file = "r-references.bib")
```

`my_citation` now contains the following text that you can use in your document: `r my_citation`

The citation style is set in the header of this document with the `csl` parameter.
The relevant references will, of course, be added to the documents references automatically.
In order for citations to work, you need to supply a .bib-file to the `bibliography` parameter in the document header.
See the [RMarkdown documentation](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html) and [Citation Style Language](http://citationstyles.org/) for further details.

## Document options
This text is set as manuscript.
If you want a thesis-like document you can change the `class` in the document header from `man` to `doc`.
You can also preview a polished journal typesetting by changing the `class` to `jou`.
Refer to the `apa6` document class [documentation](ftp://ftp.fu-berlin.de/tex/CTAN/macros/latex/contrib/apa6/apa6.pdf) for further class options, such as paper size or draft watermarks.

When creating PDF documents, line numbering can be activated by setting the `lineno` argument in the header of this document to true. This option has no effect on Word documents.

## Last words
That's all I have. Enjoy writing your manuscript. If you have any trouble or ideas for improvements, open an [issue](https://github.com/crsh/papaja/issues) on GitHub or make a pull request with the fix. ;)

# References

```{r create_r-references}
r_refs(file = "r-references.bib")
```

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
\setlength{\parskip}{8pt}
