---
title: "Getting started on new `apa_print()` methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started on new `apa_print()` methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library("papaja")
```

This document provides a brief overview of issues to consider, when implementing a new method for `apa_print()`, a convenience function to facilitate reporting of results in accordance with APA reporting guidelines.
If you consider adding a new method, read the [contributing guidelines]() first. <!-- TODO: Add link -->

## Deciding on a new method

The best way to decide on a new method to add is to solve your own problems.
If you are reporting the results of a statistical analysis that is not yet supported by `apa_print()` you probably have a good motivation and possibly prior work to build on.
If you are just looking for a way to contribute, take a look at the [open issues]() for inspiration. <!-- TODO: Add link -->

`apa_print()` is a generic, meaning it can, in principle, work on any output object with a class that is specific enough to extract the results of the analysis.
For example, objects of class `htest`, as returned by `t.test()`, `cor.test()`, `prop.test()`, etc., are named lists that follow a loose convention about the named objects they contain.

```{r htest-example}
t_test_example <- t.test(extra ~ group, data = sleep)
class(t_test_example)
str(t_test_example)
```

Hence, if we pass an `htest` object to `apa_print()` the function expects there to be named elements in the list, such as `statistic`, `estimate`, or `p.value`.
These expectations are reflected in the workings of the `apa_print.htest()` method.
Objects of less specific classes, such as `list` or `data.frame` cannot be supported, because reliable assumptions about their contents are much more limited.

## Default structure of output

Objects returned by `apa_print()` are of class `apa_results`, a named list with four elements.
The internal function `apa_print_container()` can be used to generate such a to-be-filled and -returned list:

```{r apa-results}
papaja:::apa_print_container()
```

To illustrate how `apa_results` objects are populated, let's look at the output of `apa_print.lm()`.

```{r lm-example}
# Data from Dobson (1990), p. 9.
ctl <- c(4.17, 5.58, 5.18, 6.11, 4.50, 4.61, 5.17, 4.53, 5.33, 5.14)
trt <- c(4.81, 4.17, 4.41, 3.59, 5.87, 3.83, 6.03, 4.89, 4.32, 4.69)
group <- gl(2, 10, 20, labels = c("Ctl", "Trt"))
weight <- c(ctl, trt)
lm_fit <- lm(weight ~ group)

lm_fit_apa <- apa_print(lm_fit)
```


The `estimate` element of the returned `apa_results`-list itself contains a named list with estimated parameters---in this case regression coefficients---and corresponding confidence intervals for the model.
The names of the list correspond to the names of the predictors.

```{r estimates}
lm_fit_apa$estimate
```

The `estimate`-list may contain additional elements, such as in this case the list `modelfit`, that contains estimates that quantify the quality of the model fit.

The `statistic` element of the returned `apa_results`-list again contains a named list with the same structure as `estimate`.
The two lists mainly differ with respect to their contents.
Instead of parameter estimates, `statistic` contains the corresponding inferential statistics, such as significance tests or Bayesian model comparisons.

```{r statistic}
lm_fit_apa$statistic
```

Note that the `statistics`-list misses elements for the information criteria `AIC` and `BIC`.
Because no inferential statistics on the information criteria are available, it is fine to simply drop those elements.

The `full_results` element is a named list that simply combines the results of `estimate` and `statistic` for convenience in reporting.

```{r full-result}
lm_fit_apa$full_result
```

Finally, the `table` element contains a `data.frame` of class `apa_results_table` that summarizes the results.
For more complex analyses `table` may contain a named list of `apa_result_table`s.
In essence this is simply a regular `data.frame` that follows the column naming conventions used in [`broom`](https://cran.r-project.org/package=broom) but allows for prettier printing of variable labels.

```{r table}
lm_fit_apa$table
```

Variable labels, as implemented in `papaja`, are attributes of each column that define a typeset label for the respective column.

```{r variable-labels}
letters

variable_label(letters) <- "Letters of the alphabet"
variable_label(letters)

letters
```

```{r variable-label-column}
lm_fit_apa$table$statistic
```


These variable labels are automatically used by `apa_table()` and plotting functions from the `apa_factorial_plot()`-family to create sensible default labels.
If a label is enveloped in `$` it may contain LaTeX math syntax, which is automatically converted to R expressions using [`latex2exp`](https://cran.r-project.org/package=latex2exp) for plotting.

Any new `apa_print()` method should output an object of this basic structure.

## Typesetting numeric information

`apa_results` do not contain numeric information.
Rather the numeric information has been processed for printing in accordance with APA guidelines.
There are two `papaja`-functions to facilitate the typesetting.
`printnum()` is a flexible general purpose function that wraps `formatC()` and can be used to round, set decimal as well as thousands separators, or remove leading zeros.

```{r printnum}
printnum(rnorm(3) * 1e4, digits = 3, big.mark = ".", decimal.mark = ",")
```

`printp()` is a wrapper for `printnum()` that sets appropriate defaults to report p-values in accordance with APA guidelines.

```{r printp}
printp(c(0.0001, 0.05, 0.99999))
```

Finally, `papaja` provides an internal function `papaja:::print_interval()` to typeset interval estimates.

```{r print-interval}
papaja:::print_interval(rnorm(2), conf_level = 0.95, interval_type = "CI")
```

Again, there are two wrappers that set appropriate defaults to typeset frequentist confidence and Bayesian highest density intervals.

```{r print-confint}
papaja:::print_confint(rnorm(2), conf_level = 0.95)
papaja:::print_hdint(rnorm(2), conf_level = 0.95)
```

## Typesetting model terms

When creating named lists from terms, these terms names should ideally follow tidy naming conventions.
Specifically, they should use `_` as separator, and be valid R names.
Adhering to these conventions ensures that `apa_results` can conveniently be indexed using the `$` operator.

To facilitate the generation of list names, `papaja` provides the internal function `sanitize_terms()`.

```{r sanitize-terms}
papaja:::sanitize_terms(c("(Intercept)", "Factor A", "Factor B", "Factor A:Factor B", "scale(Factor A)"), standardized = TRUE)
```

While these sanitized terms are well suited to name R objects, they are not ideal for reporting.
To facilitate typesetting term names for reporting, there is another internal function `papaja:::prettify_terms()`.

```{r prettify-terms}
papaja:::prettify_terms(c("(Intercept)", "Factor A:Factor B", "scale(Factor A)"), standardized = TRUE)
```


## Internal workflow

### Method dispatch

As with `lm` objects, it is often the case that the objects, as returned by the analysis function, may not contain all information necessary to populate the lists described above.
For example, to obtain inferential statistics it may be necessary to call `summary()`.

```{r aov-fit}
npk_aov <- aov(yield ~ block + N * P * K, npk)
npk_aov

summary(npk_aov)
```

This is why there are usually multiple `apa_print()`-methods that are called subsequently to make the function both flexible and convenient.
For convenience, `apa_print.aov()` calls `summary()` with its default arguments and passes the result onto `apa_print.summary.aov()`.

```{r apa-print-aov}
papaja:::apa_print.aov
```

This approach furthermore ensures that a variety of object types are supported while minimizing code redundancy.

### Restructuring results

The internals of `apa_print()` heavily rely on the [`broom`](https://cran.r-project.org/package=broom), a package to that provides functions turn the output of analysis functions into tidy `data.frame`s.
The objects are often processed using `broom::tidy()` and `broom::glance()` if necessary before being modified further to create the contents of the `table` element.

Once the results table has been assembled, variable labels have been assigned, and the `apa_results_table` class has been applied, `estimate`, `statistic`, and `full_result` can be derived from it.
A convenient approach is to use `apply()` rowwise.

```{r construct-estimate}
library("broom")
tidy_lm_fit <- tidy(lm_fit, conf.int = TRUE)

tidy_lm_fit$estimate <- printnum(tidy_lm_fit$estimate)
tidy_lm_fit$ci <- apply(tidy_lm_fit[, c("conf.low", "conf.high")], 1, papaja:::print_confint, conf_level = 0.95)

tidy_lm_fit[, c("estimate", "ci")]


results <- papaja:::apa_print_container()
results$estimate <- as.list(apply(tidy_lm_fit, 1, function(x) paste0("b = ", x["estimate"], " ", x["ci"])))
names(results$estimate) <- papaja:::sanitize_terms(tidy_lm_fit$term)

results
```


## User interface

A final issue to consider is to make sure that the objects users pass to `papaja` functions is appropriate.
This input validation serves to provide useful error messages to users and point them towards what is going wrong.
Currently, `papaja` relies on the internal function `validate()` for this.

```{r}
in_paren <- TRUE
papaja:::validate(in_paren, check_class = "logical", check_length = 1)
```

